Reviving the Lost Art of \\Combinator Graph Reduction

### 2023-01-30 experiments/superg/doc/paper/paper.tex
Graph reduction.

\section{Preliminaries}
\label{sec:prelim}
\textbf{TODO: write section. Currently contains only a few small fragments}
\subsection{Lambda Calculus}

\subsection{Combinatory Logic}
Starting from a lambda calculus expression, we first translate it into an expression consisting exclusively of the application of a small set of predefined functions, or \textit{combinators}.
It turns out that any lambda calculus expression can be expressed using just three combinators:
\begin{itemize}
    \item $S \ f \ g \ x = f \ x \ (g \ x)$
    \item $K \ x \ y = x$
    \item $I \ x = x$\footnote{It turns out that $SKK = I$, so even this combinator is not strictly necessary, but all reasonable implementations do include $I$, and it simplifies the example.}
\end{itemize}

For example, $\lambda x.x+1$ can be converted into the combinator expression:

\[
    S \ (S \  (K \  +) \  I) \ (K \ 1)
\]

We invite the reader to verify this by applying $x$ to this expression and reducing it to $x+1$ with the usual $\beta$-reduction rule.
While the generated expression is not as easy to read, the simplicity of its components is appealing: to execute a program compiled to combinators, an evaluator only needs to support four operations: the built-in $S$, $K$ and $I$ combinators, and function application\footnote{And any builtins of the language, like the plus operator in the example.}.
Better still, conversion to combinators removes all bound variables.
The runtime need not concern itself with passing arguments, this is all handled by the combinators.

\textbf{TODO: } Combinators $B$ and $C$.

\subsection{Bracket Abstraction}

\subsection{Graph Reduction}
Must cover: outermost reduction.

@@

\begin{figure*}
    \includegraphics[width=2\columnwidth]{tigre_layout}
    \centering
    \caption{
        Memory layout of a node in the TIGRE-style engine.
        Each square represents a byte of memory.
    }
    \label{fig:tigre_layout}
\end{figure*}

>>>>

\section{Preliminaries}
\label{sec:prelim}
\subsection{Lambda Calculus}
For an accessible introduction to the Lambda Calculus, we can recommend~\cite{rojas_tutorial_2015}.
Most introductory resources, including~\cite{rojas_tutorial_2015}, present Lambda Calculus with named bindings.
Kiselyov's algorithms~\cite{kiselyov_lambda_2018} use De Bruijn indices~\cite{de_bruijn_lambda_1972} instead.

%The lambda ($\lambda$) calculus is perhaps the smallest programming language.
%At its core there is only the definition of functions (\emph{abstraction}) and the \emph{application} of a function to an argument.
%We use the $\lambda$ symbol to denote abstraction, e.g. $\lambda x.x$ defines a function (unnamed) that takes an argument $x$ (the $x$ before the dot), and when applied to an $y$ returns that same value $y$.
%This application may be written as $(\lambda x.x) y$.
%$\lambda x.x$ is in fact the definition of the \emph{identity} function: a function that returns its argument unchanged.
%we can nest abstractions to create functions that accept multiple arguments, for example $\lambda x.\lambda y. x$, a function that takes two arguments and returns the first (the second is discarded).
%
%Functions may be arguments to other functions: $(\lambda x.x) (\lambda x.x)$ is a valid expression in the lambda calculus (it evaluates to $\lambda x.x$).
%
%TODO: Outermost reduction
%\url{https://personal.utdallas.edu/~gupta/courses/apl/lambda.pdf}
%
\subsection{Combinatory Logic}
Chapter 16 of~\cite{spj_impl} starts with a good introduction to Combinatory Logic and the Bracket Abstraction algorithm.

%Starting from a lambda calculus expression, we first translate it into an expression consisting exclusively of the application of a small set of predefined functions, or \textit{combinators}.
%It turns out that any lambda calculus expression can be expressed using just three combinators:
%\begin{itemize}
%    \item $S \ f \ g \ x = f \ x \ (g \ x)$
%    \item $K \ x \ y = x$
%    \item $I \ x = x$\footnote{It turns out that $SKK = I$, so even this combinator is not strictly necessary, but all reasonable implementations do include $I$, and it simplifies the example.}
%\end{itemize}
%
%For example, $\lambda x.x+1$ can be converted into the combinator expression:
%
%\[
%    S \ (S \  (K \  +) \  I) \ (K \ 1)
%\]
%
%We invite the reader to verify this by applying $x$ to this expression and reducing it to $x+1$ with the usual $\beta$-reduction rule.
%While the generated expression is not as easy to read, the simplicity of its components is appealing: to execute a program compiled to combinators, an evaluator only needs to support four operations: the built-in $S$, $K$ and $I$ combinators, and function application\footnote{And any builtins of the language, like the plus operator in the example.}.
%Better still, conversion to combinators removes all bound variables.
%The runtime need not concern itself with passing arguments, this is all handled by the combinators.
%
%\textbf{TODO: } Combinators $B$ and $C$.
%
\subsection{Bracket Abstraction}

\subsection{Graph Reduction}
A good tutorial-style introduction to Graph Reduction may be found at~\cite{peyton_jones_implementing_2000}.
For a more formal and detailed reference, see~\cite{spj_impl}.

Graph Reduction is a common technique for implementing non-strict functional programming languages, i.e. languages that evaluate expressions only when their value is needed.
Nodes in the graph are either constant values (leaf nodes) or application nodes (representing the application of a function to an argument).
Edges are used to connect application nodes with their function and argument nodes.
The usual convention is to denote application nodes with \emph{@}.
Here is how we might represent the expression \texttt{square 3}:

\begin{verbatim}
       @
      / \
square   3
\end{verbatim}

This graph has three nodes: a reference to the function \texttt{square}, an argument 3, and an application node.
We can \emph{reduce} this graph by applying rewrite rules to it. For example, we know that an expression \texttt{square x} produces a value $x*x$.
For the above graph, we can therefore state that:

\begin{verbatim}
       @
      / \    reduces to    9
square   3
\end{verbatim}

The example graph happens to be a tree, but this is not always the case: A node may be referenced by multiple application nodes.
Consider the graph for \texttt{square (square 3)}:

\begin{verbatim}
       @
      / \
     |   @
     |  / \
  square   3
\end{verbatim}

The two application nodes \emph{share} the \texttt{square} node.
When we perform the reduction of \texttt{square 3}, the upper application node maintains its (now exclusive) reference to \texttt{square}, and the other application node is replaced with the constant value 9.
Sharing nodes is important for efficiency because it prevents unnecessary re-evaluation of expressions.
The graph representation of \texttt{(square 3) + (square 3)} highlights this:

\begin{verbatim}
      @ 
     / \
    @  |
   / \ |
  +   \|
       @
      / \
square   3
\end{verbatim}

We would like to evaluate \texttt{square 3} only once.
One way to do this is to replace the top application node of an evaluated expression with the new value.
For the example graph above, we get:

\begin{verbatim}
      @ 
     / \
    @  |
   / \ |
  +   \|
       9
\end{verbatim}

Now both the left and right sides of the + see the updated value (notice that we are passing the two arguments to + using a double application node).

A graph could have multiple possible rewrites (reduction steps) that we could perform, in which case we need to decide what rewrite rules should be applied first.
There are many possible ways to make this decision~\cite{spj_impl}, but most practical implementations choose a \emph{left-outermost} reduction strategy.
In the graph representation, this amounts to traversing the graph from the root, following \emph{left} edges of application nodes.
This chain of edges from the root down is called the \emph{spine} of the reduction.
Eventually, we find a leaf node and apply the rewrite rule that matches that particular node (if any apply).
The reduction then repeats from the root again, until eventually, no more rewrites are possible.

At this point, it could be that the graph has been reduced to a constant value (e.g. 9).
Another possibility is parts of the graph are still reduceable, but our reduction strategy does not `reach' it.
For example, we can have the following graph (including a $K$ combinator) that our strategy will not reduce:

\begin{verbatim}
     @
    / \
   K   @
      / \
square   3
\end{verbatim}

$K$ requires two arguments, but we have given it only one, so the usual rewrite rule $K\ x\ y \Rightarrow x$ does not apply.
Because the graph contains a subexpression that \emph{could} be reduced, we say the reduction has reached \emph{Weak Head Normal Form}, rather than \emph{Normal Form}.

@@

\begin{figure*}
    \includegraphics[width=1.4\columnwidth]{tigre_layout}
    \centering
    \caption{
        Memory layout of a node in the TIGRE-style engine.
        Each square represents a byte of memory.
    }
    \label{fig:tigre_layout}
\end{figure*}
